---
title: "Reproducible Research: Peer Assessment 1"
author: Mick Hittesdorf
date: "11/16/2014"
output: html_document
keep_md: true
---


## Loading and preprocessing the data
```{r}
require(plyr)
require(lattice)

data <- read.csv("activity.csv")
summary(data)
```

## What is mean total number of steps taken per day?
```{r histogram}
require(plyr)
total_steps_per_day <- ddply(data, .(date), summarise, total_steps = sum(steps))
with(total_steps_per_day, hist(total_steps, main="Total Steps Per Day"))
```

#### Calculate and report the mean and median total steps per day

```{r}
mean_total_steps_per_day <- sprintf("%.2f", with(total_steps_per_day, mean(total_steps, na.rm = TRUE)))
median_total_steps_per_day <- sprintf("%.2f", with(total_steps_per_day, median(total_steps, na.rm = TRUE)))

```
```
Mean total steps per day: `r mean_total_steps_per_day`
Median total steps per day: `r median_total_steps_per_day`
```

## What is the average daily activity pattern?

#### Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
mean_steps_per_interval <- ddply(data, .(interval), summarise, mean_steps_per_interval = mean(steps, na.rm=TRUE))
with(mean_steps_per_interval, plot(interval, mean_steps_per_interval, type='l', main = "Mean steps per interval averaged over each day"))

```

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
max_steps_per_interval_index <- with(mean_steps_per_interval, which(mean_steps_per_interval == max(mean_steps_per_interval)))
max_interval_number <- mean_steps_per_interval[max_steps_per_interval_index,1]
```

Interval with the maximum average daily number of steps: `r max_interval_number`

## Imputing missing values

#### Calculate and report the total number of missing values in the dataset (i.e .the total number of rows with NA's)
```{r}
total_number_missing_values <- sum(with(data, is.na(steps)))
```
Total number of missing values (NA's): `r total_number_missing_values`

#### Create a new dataset that is equal to the original but with the missing data filled in
lookup_mean_steps_per_interval <- function(interval) {
  mean_steps_for_interval <- mean_steps_per_interval[mean_steps_per_interval$interval == interval, 2]
  return(mean_steps_for_interval)
}

```{r}
merged <- merge(data, mean_steps_per_interval)
ordered <- with(merged, merged[order(date),])
ordered_na_indices <- which(is.na(ordered$steps))
interval_means <- ordered[ordered_na_indices,"mean_steps_per_interval"]
na_indices <- with(data,which(is.na(steps)))
data_no_nas <- data
data_no_nas$steps[na_indices] <- ordered$mean_steps_per_interval[na_indices]
```

#### Plot a histogram again after missing data has been imputed

```{r}
total_steps_per_day_no_nas <- ddply(data_no_nas, .(date), summarise, total_steps = sum(steps))
with(total_steps_per_day_no_nas, hist(total_steps, main="Total Steps Per Day"))

```

#### Recalculate mean and median after missing values have been imputed
```{r}
mean_total_steps_per_day_no_nas <- sprintf("%.2f", with(total_steps_per_day_no_nas, mean(total_steps, na.rm = TRUE)))
median_total_steps_per_day_no_nas <- sprintf("%.2f", with(total_steps_per_day_no_nas, median(total_steps, na.rm = TRUE)))

```

```
Mean total steps per day: `r mean_total_steps_per_day_no_nas`
Median total steps per day: `r median_total_steps_per_day_no_nas`
```

#### Do these values differ from original calculations? 

Yes. The mean and median are now equal!

## Are there differences in activity patterns between weekdays and weekends?
```{r}
weekday_or_weekend <- function(date_factor) {
  result <- c()
  if (is.factor(date_factor)) {
    date <- as.Date(date_factor)
    days <- weekdays(date)
    weekend_days <- c("Saturday", "Sunday")
    result <- vector(mode="character", length(days))
    i = 1
    for (day in days) {
      if (day %in% weekend_days) {
        result[i] = "weekend"
      } else {
        result[i] = "weekday"
      }
      i = i + 1
    }
  }
  return(result)
}

data_no_nas_with_factors <- transform(data_no_nas, day_type = weekday_or_weekend(date))

mean_steps_per_interval_day_types <- ddply(data_no_nas_with_factors, .(interval, day_type), summarise, mean_steps_per_interval = mean(steps, na.rm=TRUE))
with(mean_steps_per_interval_day_types, xyplot(mean_steps_per_interval ~ interval | day_type, layout = c(1,2), type="l"))


```
```
